Gaia identifies VFTS682 as the most massive runaway star known to date

### 2018-05-12 VFTS682.tex
text done to end

Using the 651 stars selected as described in \Secref{data:gaia} (a
subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
$\langle\mu_\mathrm{RA}\rangle = 1.7240\pm0.0002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DEC}\rangle =0.6986\pm0.0003\,\mathrm{mas\ yr^{-1}}$. We note that these values are
in good agreement with what is found by \cite{lennon:18}. Subtracting these values from the
proper motion of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\mu_\mathrm{RA} = 0.12\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu_\mathrm{DEC} =
0.08\pm 0.08\,\mathrm{mas\ yr^{-1}}$. We note that the error budget is
dominated by the errors on the proper motion components of VFTS682.

@@

\section{Discussion}
\label{sec:discussion}
\todo{Selma: Should add caveat that it may be a spurious detection.  That reconfirmation in following epochs is warranted. }

@@

Dynamical ejections due to N-body interactions typically (although, not necessarily) eject the least
massive star among those interacting \todo{ref}. This means that, just
based on the kinematic properties of VFTS682, we would expect several
stars with initial masses larger than $\sim$$150\,M_\odot$ in the
cluster R136.
This is consistent with the detection
of extremely massive stars by \cite{crowther:10} in the core of the
cluster. The projected rotational equatorial
velocity\footnote{However, the determination of the rotational
  velocity for stars showing lines in emission is complicated because
  of the optically thick wind screening the surface of the star, and should be
  considered with caution.} of VFTS682
reported by \cite{schneider:18} is $v\sin(i)<200\,\kms$, which is in
line with the average rotation rate of massive stars in the region
\citep[][]{ramirez-agudelo:15}. This suggests that VFTS682 (i) has not
experienced rotationally induced chemically homogeneous evolution
\citep[][]{maeder:00,demink:09,marchant:16}, and (ii) it has not
accreted mass from a binary companion, nor will it, since the multi-epoch
data of the VFTS survey rule out the presence of a close companion at the
present day. Moreover, the spectral type of VFTS682
\citep[WNh5,][]{bestenlehner:11} is the same as R136a1-a3, i.e.~the
three most massive stars detected in the core of the cluster by
\cite{crowther:10}, with an astonishing similarity in particular with
the spectrum of R136a3. Therefore,
VFTS682 might be an ideal target to constrain the stellar physics of
stars with masses well above $\sim$$100\,M_\odot$: its isolation makes
it an easier target for observations compared to the similar stars
present in the crowded core of R136.

The similarities between VFTS682 and the WNh5 stars in the core of
R136 are also in agreement with the ``bully binary'' model of
\cite{fujii:11}. Based on their numerical results, they suggested that
early in the evolution of a cluster, dynamical interactions form an extremely
massive binary, which then tightens its orbit by ejecting other stars passing
by. Interpreting our results for VFTS682 through the lens of their simulations
suggests the presence of a binary with total mass
$M_1+M_2\gtrsim 300\,M_\odot$ in the core of the cluster. Such bully
binary could be R145 according to \cite{fujii:11}.

The kinematic age of VFTS682 puts an
upperlimit to the timescale to form such a ``bully binary'' in
R136. The cluster must have been at the very beginning of its
evolution, given the age estimate of $\lesssim 2$\,Myr
\citep[][]{crowther:10,sabbi:12} and the kinematic age of VFTS16. If the
cluster is indeed younger than the shortest stellar lifetime
\citep[$\sim$3\,Myr, e.g.,][]{zapartas:17}, then the alternative
explanation for ejection of VFTS682 from a binary
system is excluded since the region is too young for stars
to have experience core-collapse already.

\todo{check \cite{banerjee:12} and rewrite next paragraph, they
  have stuff} Unfortunately, their simulations did not include stars more
massive than $100\,M_\odot$, so it is difficult to predict, based on
the existence of VFTS682, how many other stars should have been
ejected from the cluster already and what their mass
distribution is. A comprehensive study of the kinematic properties of the
large sample of stars with extremely large inferred masses visible
around R136 is encouraged.

\cite{lennon:18} have carried out a similar study on VFTS16, another
massive star ($\sim90\,M_\odot$) in the 30 Doradus region, previously
known to be a runaway from the value of
its radial velocity. They also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster. 
The value of $\tau_\mathrm{kin}\simeq0.85$\,Myr we find for VFTS682 (see \Secref{sec:r136_origin}) is smaller
than the corresponding value for VFTS16: \cite{lennon:18} inferred a kinematic age of
$\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.

\begin{itemize}
\item is R136 a single young cluster or a merger
\item estimate the influence of the gravitational potential of R136,
  what is its total mass and relaxation time?
\item Make link to dynamical interaction proposed to form BHBH
  systems.
  \item from Fabian's email (see comment in tex):% One last aspect: I am always wondering what happens to the star/binary that ejected 682. There is a huge linear momentum in this star of ~150 Msun * 60 km/s. Even if the bully was a 300 Msun total object, it would obtain a kick of 30 km/s. Is this enough to eject the bully star? This velocity would already be 100 km/s for a 90 Msun bully (I was thinking of 016 here... 'Cause I do not know what ejected this object either). Of course, if there is a third body, these velocities are no longer correct. The escape velocity from R136 at 0.5pc for a total mass of 10^5 Msun would only be of order 20 km/s... So a 300 Msun bully is at the edge of what is possible and one might need something even more massive or another very massive object will also be ejected. Or am I wrong here? Do the Banerjee and Co simulations say something about this?
\end{itemize}

% \todo{Merge Selma's my\_bib bibtex file with Mathieu's and prevent
% double refs. } %% copied everything in one file

>>>>

Using the stars selected as described in \Secref{data:gaia} (a
subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
$\langle\mu_\mathrm{RA}\rangle = 1.7240\pm0.0002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DEC}\rangle = 0.6986\pm0.0003\,\mathrm{mas\
  yr^{-1}}$. These values are
in good agreement with what is found by \cite{lennon:18}. However, we
emphasize their sensitivity to the choice of stars adopted to define
the reference frame. Subtracting these values from the
proper motion of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\delta\mu_\mathrm{RA} = 0.12\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\delta\mu_\mathrm{DEC} =
0.08\pm 0.08\,\mathrm{mas\ yr^{-1}}$. The error budget is
dominated by the errors on the proper motion components of VFTS682.

@@

\section{Discussion}
\label{sec:discussion}
\todo{subsection on validation with HST data?Discussion on
  correlations between pmRA and pmDEC?Not sure where to start. HST
  data for VFTS682 dont seem to be public.}
\todo{Selma: Should add caveat that it may be a spurious detection.  That reconfirmation in following epochs is warranted. }

@@

Dynamical ejections due to N-body interactions typically (although, not necessarily) eject the least
massive star among those interacting \cite[e.g.,][]{banerjee:12}. This means that, just
based on the kinematic properties of VFTS682, we would expect several
stars with initial masses larger than $\sim$$150\,M_\odot$ in the
cluster R136.
This is consistent with the detection
of extremely massive stars by \cite{crowther:10} in the core of the
cluster. The projected rotational equatorial
velocity\footnote{However, the determination of the rotational
  velocity for stars showing lines in emission is complicated because
  of the optically thick wind screening the surface of the star, and should be
  considered with caution.} of VFTS682
reported by \cite{schneider:18} is $v\sin(i)<200\,\kms$, which is in
line with the average rotation rate of massive stars in the region
\citep[][]{ramirez-agudelo:15}. This suggests that VFTS682 (i) has not
experienced rotationally induced chemically homogeneous evolution
\citep[][]{maeder:00,demink:09,marchant:16}, and (ii) it has not
accreted mass from or merged with a binary companion, nor will it, since the multi-epoch
data of the VFTS survey rule out the presence of a close companion at the
present day. Moreover, the spectral type of VFTS682
\citep[WNh5,][]{bestenlehner:11} is the same as R136a1-a3, i.e.~the
three most massive stars detected in the core of the cluster by
\cite{crowther:10}, with an astonishing similarity in particular with
the spectrum of R136a3. Therefore,
VFTS682 might be an ideal target to constrain the stellar physics of
stars with masses well above $\sim$$100\,M_\odot$: its isolation makes
it an easier target for observations compared to the similar stars
present in the crowded core of R136.

\citet{banerjee:12} used N-body simulations of fully segregated
clusters with all massive stars in binaries to suggest that VFTS682
was ejected from R136, in agreement with our findings. They
demonstrated that the cluster potential does not significantly change
the velocity of the star after the ejection. In their
model, they relied on (dynamically driven) stellar mergers to explain the high masses of
VFTS682 and the massive members of R136. While it is not possible to
robustly exclude that VFTS682 is itself a merger, as already
mentioned, its spectrum does not show any of the obvious signatures
like fast rotation. To eject such massive object, the cluster must have produced
a large number of massive runaways. Indeed, several relatively
isolated massive stars are observed in the region, some with known
large radial velocities. A comprehensive study of the kinematic
properties of all the massive stars surrounding R136 might shed light
on whether some can be unequivocally identified as merger products. It
is also possible that the star or binary that caused the ejection of
VFTS682 might have been ejected in the opposite direction, and is also
in isolation at present day.

The similarities between VFTS682 and the WNh5 stars in the core of
R136 are also in agreement with the ``bully binary'' model of
\cite{fujii:11}. Based on their numerical results, they suggested that
early in the evolution of a cluster, dynamical interactions form an extremely
massive binary, which then tightens its orbit by ejecting other stars passing
by. Interpreting our results for VFTS682 through the lens of their simulations
suggests the presence of a close binary with total mass
$M_1+M_2\gtrsim 300\,M_\odot$ in the core of the cluster. Such bully
binary could be R145 according to \cite{fujii:11}, and it might be an
ideal observational candidate for a dynamically formed progenitor system of
a binary black-hole, provided that stars this massive can avoid a
pair-instability supernova \cite[][]{fowler:64, rakavy:67} at LMC
metallicity \citep[see also][]{langer:07}. Similarly, the final fate of VFTS682 could be either a
pair-instability supernova without compact remnant formation, or
possibly direct collapse to a black hole above the $2^\mathrm{nd}$
mass gap. The amount of mass loss of these stars will determine their final core
mass and thus their final fate.

The kinematic age of VFTS682 puts an
upperlimit to the timescale to form the ``bully binary'' in
R136. The cluster must have been at the very beginning of its
evolution, given the age estimate of $\lesssim 2$\,Myr
\citep[][]{crowther:10,sabbi:12} and the kinematic age of VFTS682. If the
cluster is indeed younger than the shortest stellar lifetime
\citep[$\sim$3\,Myr, e.g.,][]{zapartas:17}, then the alternative
explanation for ejection of VFTS682 from a binary
system is excluded since the region is too young for stars
to have experience core-collapse already.

The variability of VFTS682, reminiscent of LBV stars, also suggests
that VFTS682 (and therefore its analogs in the core of R136) might
experience enhanced mass loss episodes in LBV eruptions. \citet{smith:15} made the highly
debated\footnote{See, e.g., \cite{humphreys:16, davidson:16, smith:16}.}
claim that LBV stars are typically isolated form O-type stars. The fact that VFTS682 is a dynamically
ejected runaway which might evolve into an LBV star suggests that
dynamical ejections can also play a role in explaining the apparent
isolation of at least some LBV stars.

\citet{lennon:18} carried out a study similar to ours on VFTS16, another
massive star ($\sim$$90\,M_\odot$) in the 30 Doradus region, previously
known to be a runaway from the value of
its radial velocity. They also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster. 
The value of $\tau_\mathrm{kin}\simeq0.85$\,Myr we find for VFTS682 (see \Secref{sec:r136_origin}) is smaller
than the corresponding value for VFTS16: \cite{lennon:18} inferred a kinematic age of
$\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.

The numerical simulations from \cite{oh:16} suggest that dynamical
interaction eject the majority of the stars during of shortly after the cluster
core-collapse. The large number of isolated massive stars around it
suggest that R136 has already evolved past the
time of maximum stellar density. This might have implications for the
question on whether the cluster formed via a monolithic collapse, or
as a (potentially ongoing) merger of several sub-clusters.

\citet{oh:16} also showed that the mass
function of the ejected star depends on the cluster initial conditions
(whether it is segregated, what is the binary fraction, what period
distribution is assumed for the initial binaries), therefore studies on
the population of isolated massive stars in the surroundings of R136
might shed light on its initial stellar population and dynamical
state.

VFTS682 is the most massive runaway known to date, and its ejection
from the cluster R136 implies that it is only the ``tip of the
iceberg'' of possibly extremely massive runaways in the
region. Studies of this population, enabled by recent Hubble Space
Telescope and Gaia observations will put constraints on the evolution
of these extreme stars, together with the formation and evolution of
the central cluster itself.
